---
title: "Some Examples of Nonparametric Combination"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Vignette Info

These examples come from the textbook section 4.6.

```{r package}
library(permuter)
source("../R/t2p.r")
```

## Massaro-Blair

To compare the numbers of breeding yellow-eyed penguin pairs from the two groups, the authors performed a bootstrap test using the raw difference of sample means as a test statistic and obtained a significant result (p = 0.009). However, in this study, there was not only an empirical difference between the two means, but also between the standard deviations, since the variance was found to be much smaller on Stewart Island. Here we wish to analyze the same data using the NPC methodology (instead of the bootstrap test) and applying the MA procedure, in order to jointly evaluate location ($T_\mu^*$) and scatter ($T_{\sigma^2}^*$) aspects, which are supposed to be responsible for the difference betweenthe two groups. The null hypothesis $H_0: Y_1 =d Y_2$ implies the event $\{E[Y_1] = E[Y_2]\} \cap \{E[Y_1^2] = E[Y_2^2]\}$. In particular, we have examined the location-aspect by means of a standard t-test over raw data and the scatter-aspect using transformed data (Y is second moments). Note
that this solution is exact because the null hypothesis states the irrelevance of feral cats, and so the permutation testing principle applies.

This exerpt comes from p 238.

```{r massaro-blair}
data(massaro_blair)
B = 500
T = array(0,dim=c((B+1),2))
ID = massaro_blair[,"group"]
Y = massaro_blair[,"Y"]

T[1,1] = mean(Y[ID==1])-mean(Y[ID==2])
T[1,2] = mean(Y[ID==1]^2)-mean(Y[ID==2]^2)

for(bb in 2:(B+1)){
  y.star=sample(Y)
  T[bb,1] = mean(y.star[ID==1])-mean(y.star[ID==2])
  T[bb,2] = mean(y.star[ID==1]^2)-mean(y.star[ID==2]^2)
}

P = t2p_old(abs(T))
P[1,]
```

## Flies

The fly example appears in the vignette for examples in chapters 1-4. (see p 253)