---
title: "NSGK"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## NSGK

summary of the vignette

### Set up for analysis

```{r setup}
library(permuter)
library(dplyr)
library(ggplot2)

data(nsgk)
time_stamps <- c(36, 32, 35, 37, 31, 35, 40, 32)
nvideos <- length(unique(nsgk$video))
ncategories <- length(unique(nsgk$domain))
nraters <- max(nsgk$rater)

head(nsgk)
```
We have to turn this table into the right data format: for each tag-video pair, we must create a matrix of 0s and 1s, where 1 in the $i, j$ position indicates that rater $i$ saw the tag at timestamp $j$.  We'll have 8 matrices for each tag, for a total of \Sexpr{ncategories*nvideos} matrices. We'll store these matrices in a list of lists: the first level of the list indexes the tag, and for each tag there is a list of 8 matrices, one for each video.

```{r table_to_matrices}
unique_tags <- sort(unique(nsgk$domain))
nsgk_mat <- list()
for(tag in unique_tags){
  domain <- list()
  for(v in seq_len(nvideos)){
    thisvideo <- nsgk %>% 
                filter(domain == tag) %>%
                filter(video == v)
    tag_mat <- matrix(0, nrow = nraters, ncol = time_stamps[v])
    for(row in seq_len(nrow(thisvideo))){
      tag_mat[thisvideo[row, "rater"], thisvideo[row, "time_stamp"]] <- 1
    }
    domain[[v]] <- tag_mat
  }
  nsgk_mat[[length(nsgk_mat) + 1]] <- domain
}
names(nsgk_mat) <- unique_tags

length(nsgk_mat)
length(nsgk_mat[[1]])
dim(nsgk_mat[[1]][[1]])

```

### Summary statistics per video

We compute the mean and SD of the number of times a tag was observed in each video.
Rows index the tag and columns index the video. 
```{r summary_stats}
video_means <- t(sapply(nsgk_mat[1:20], function(tt){
                        sapply(tt, mean)
                        }))
head(video_means)

video_sds <- t(sapply(nsgk_mat[1:20], function(tt){
                        sapply(tt, sd)
                        }))
head(video_sds)

```

### Computing inter-rater reliability


```{r irr_results, cache=TRUE}
npc_results <- lapply(nsgk_mat[1:20], function(tt){
  video_specific_results <- lapply(tt, function(x){
                                        res <- irr_ts_distribution(x, 
                                          reps = 1000,    
                                          keep_dist = TRUE, seed = 101)
                                  })
  tag_distribution <- sapply(video_specific_results, function(x) x$dist)
  tag_pvalues      <- sapply(video_specific_results, function(x) x$pvalue)
  tag_concordance  <- sapply(video_specific_results, function(x) x$obs_ts)
  tag_npc_res <- irr_npc_distribution(tag_distribution, size = time_stamps, pvalues = tag_pvalues)
  return(list("pvalues_per_video" = tag_pvalues,
              "concordance_per_video" = tag_concordance,
              "tag_global_pvalue" = tag_npc_res$pvalue))
})

head(npc_results, 3)
```

Let's put it all together:
```{r results_to_csv, cache=TRUE}
ncategories <- 20
nsgk_res <- matrix(nrow = ncategories, ncol = (1 + 1 + 8*4))
for(i in seq_len(ncategories)){
  nsgk_res[i, ] <- c(unique_tags[i],
                     npc_results[[i]]$tag_global_pvalue,
                     npc_results[[i]]$concordance_per_video,
                     npc_results[[i]]$pvalues_per_video,
                     video_means[i,],
                     video_sds[i,])
}
colnames(nsgk_res) <- c("tag_index",
                        "global_pvalue",
                        paste("video", 1:8, "concordance", sep = ""),
                        paste("video", 1:8, "pvalue", sep = ""),
                        paste("video", 1:8, "mean", sep = ""),
                        paste("video", 1:8, "sd", sep = ""))
head(nsgk_res)
#write.csv(nsgk_res, file = "nsgk_results.csv", row.names=FALSE)
```

Overall, what's the relationship between concordance and P?
```{r plot_pvalues, fig.width=8}

nsgk_res %>%
  as.data.frame() %>%
  mutate(avg_concordance = 
               apply(nsgk_res[, 3:10], 1, mean)) %>%
  ggplot(aes(x = avg_concordance, y = global_pvalue)) + 
  geom_point(color = "#629e1f", alpha = 0.6, size = 3) +
  xlab("Average Concordance") +
  ylab("Overall P-value") +
  ggtitle("Average Concordance Across 8 Videos vs P-value") +
  theme(
    panel.background = element_rect(fill = "#E8EBEF"),
    axis.text = element_text(size = 16, color = "#143264"),
    axis.title = element_text(size = 18, color = "#143264"),
    title = element_text(color = "#143264", size = 18)
  )
```